Retrofitting Effect Handlers onto OCaml

### 2020-11-17 retro-concurrency.tex
edits

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. Since C allows pointers into the stack, care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

Similar to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
implementing continuations, but our continuations are one-shot. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies. They best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines as a macro-expressible translation, and present an embedding in Lua
and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
uses its own stack similar to how each handled computation runs in its own
fiber in Multicore OCaml. Similar to Multicore OCaml, Lua does not allow
coroutines to include C frames. It is useful to note that Lua is an interpreted
language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.

>>>>

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. Since C allows pointers into the stack, care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

Similarly to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
implementing continuations, but our continuations are one-shot. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines as a macro-expressible translation, and present an embedding in Lua
and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
uses its own stack similar to how each handled computation runs in its own
fiber in Multicore OCaml. Lua does not allow coroutines to include C frames, as in
Multicore OCaml. It is useful to note that Lua is an interpreted
language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.
